Paper Analysis

### 2025-04-17 Paper_WriteUp/Paper_Plots.qmd
put PPD code in for Magnus & Nihal, I think it's currently bugged though

```{r figure 7 - PPDs for Magnus & Nihal}

>>>>

```{r figure 7 - PPDs for Magnus & Nihal}
#where models are stored
model_path = here("results_revision", "lichessGrandmasters")

#read in the model and users
model = readRDS(file = here(model_path, "all_rated_bullet_model_prev_n1.RDS"))
users = readRDS(file = here(model_path, "users_bullet.RDS"))

#grandmaster data
data_path = paste0(all_data_path[4], "/")
files = list.files(data_path)
lichess_data = files %>%
  map_dfr(~read_player(data_path, .x))

#Tidy data

#selecting bullet 60+0 only
bullet_60 = lichess_data  %>%
  mutate(Event = tolower(Event)) %>%
  filter(Event == "rated bullet game", Variant == "Standard", TimeControl == "60+0") %>%
  dplyr::select(c("White", "Black", "Result", "UTCDate", "UTCTime", "Username", "WhiteElo", "BlackElo"))

#remove the duplicate rows
bullet_60_no_dup = bullet_60[!duplicated(bullet_60),]

#final 1000 games each user plays
last_games = bullet_60_no_dup %>%
  filter(Username %in% users) %>%  
  group_by(Username) %>%
  arrange(UTCDate, UTCTime, .by_group = TRUE) %>%
  slice_tail(n = 1000) %>%
  ungroup()

#only consider previous game for history
tidy_games = map_dfr(users, get_hist, last_games, prev_n = 1) |> 
  as_tibble()

#adding performance history
init_data = tidy_games |>
  mutate(WhiteElo = as.numeric(WhiteElo),
         BlackElo = as.numeric(BlackElo),
         focal_user = ifelse(focal_white == 1, White, Black),
         elo_diff = ifelse(focal_white == 1,
                           WhiteElo - BlackElo, BlackElo - WhiteElo),
         focal_id = match(focal_user, users),
         focal_rating = ifelse(focal_white == 1, WhiteElo, BlackElo),
         opp = ifelse(focal_white == 1, Black, White),
         opp_rating = ifelse(focal_white == 1, BlackElo, WhiteElo),
         UTCDateTime = ymd_hms(paste0(UTCDate, "_", UTCTime))) |>
  dplyr::select(focal_user, focal_id, focal_white, WhiteElo, BlackElo,
                focal_win_prop, elo_diff, focal_result, opp, opp_rating,
                focal_rating, UTCDateTime) |>
  group_by(focal_id) |>
  mutate(time_diff = UTCDateTime - lag(UTCDateTime, default = NA), 
         ave_prop = ifelse(time_diff <= 300,  #if games played in diff sessions, make history same as draw (ie NULL)
                           lag(focal_win_prop, default = 0.5), #removing standardization
                           0.5)) |>
  ungroup() |>
  filter(focal_result != 0.5)

#ids of players to plot
magnus_id = which(users == "DrNykterstein") #magnus
nihal_id = which(users == "nihalsarin2004") #nihal

#choose who to plot
focal_player_used = magnus_id

#slightly less than 1000 games since filtered out draws
focal_init_data = init_data |>
  filter(focal_user == users[focal_player_used])

focal_init = data.frame("Player" = users[focal_player_used],
                        "Rating" = ifelse(focal_init_data$focal_white == 1, 
                                          focal_init_data$WhiteElo, focal_init_data$BlackElo))

num_sims = 100 #number of rating curve simulations over final games
num_games = nrow(focal_init_data) #number of games to predict over

#parameter draws for focal player
focal_draws = model$draws() |> as_draws_df() |>
  dplyr::select(starts_with(c("beta[", "gamma"))) |>
  mutate(draw = row_number()) |> 
  pivot_longer(cols = "beta[1,1]":paste0("beta[2,", as.character(length(users)), "]")) |> 
  mutate(param = stringr::str_extract(name, pattern = "\\d"),
         id = stringr::str_extract(name, pattern = "\\d+]"),
         id = stringr::str_replace(id, "\\]", ""),
         player_id = paste0("beta[", id, "]")) |> 
  dplyr::select(-name) |> 
  pivot_wider(names_from = param, values_from = value) |> 
  rename(beta1 = `1`, beta2 = `2`) %>%
  filter(id == focal_player_used) %>%
  filter(draw %in% 1:num_sims)

## the covariates for the future games
future_games_cov = tibble(focal = as.character(focal_init_data$focal_id),  
                          color = focal_init_data$focal_white,
                          elo_diff = focal_init_data$elo_diff,
                          hist = focal_init_data$ave_prop,
                          focal_rating = focal_init_data$focal_rating,
                          opponent = focal_init_data$opp,
                          opponent_rating = focal_init_data$opp_rating)

#storage
glicko_sim = matrix(NA, nrow = num_games, ncol = num_sims)
game_sim = matrix(NA, nrow = num_games, ncol = num_sims)
#probs = matrix(NA, nrow = num_games, ncol = num_sims) #for debugging

#sim
set.seed(2025)
for (i in 1:num_games) {
  #current game info
  curr_game_info = future_games_cov[i,]

  #indicator for colour effect, 1 is white, 0 is black
  colour_ind = curr_game_info$color

  #elo diff for rating effect
  rating_diff = curr_game_info$elo_diff

  #opponent name and rating
  opp = curr_game_info$opponent
  opp_rating = curr_game_info$opponent_rating

  #simulate wins/loss on current game
  for (j in 1:num_sims) {
    curr_draw = focal_draws %>% filter(draw == j)

  ## if first game use default history (0), otherwise use the previous
  if (i == 1) {
    curr_rating = curr_game_info$focal_rating
    curr_hist = 0.5 #assume they're current past performance is at their usual level
  } else {
    curr_hist = curr_game_info$hist
    curr_rating = glicko_sim[i - 1, j]
  }

  #find probability
  lin_comb = curr_draw$beta1 + (curr_draw$beta2 * curr_hist) +
             (curr_draw$gamma1 * colour_ind) +
             (curr_draw$gamma2 * rating_diff)
  prob = invlogit(lin_comb)

  #simulate result
  sim_result = sample(c(1, 0), size = 1, prob = c(prob, 1 - prob))

  #find rating
  games = data.frame(Week = i,
                     Player1 = users[focal_player_used],
                     Player2 = opp,
                     Score = sim_result)
  initstate = tibble(Player = c(users[focal_player_used], opp),
                     Rating = c(curr_rating, opp_elo),
                     Deviation = 50,
                     Volatility = 0.04)

  sim_glick = glicko2(games, status = initstate, history = TRUE, gamma = colour_ind)
  ## extract the rating of focal player and store it
  focal_glick = sim_glick$history[users[focal_player_used], ,]

  #store current result and rating
  glicko_sim[i, j] = as.numeric(focal_glick)[1]
  game_sim[i, j] = sim_result
  }
  if(i %% 10 == 0) {print(i)} #progress of number of games simulated
}

#true glicko rating of all games before last 1000
true_glicko_first = bullet_60_no_dup |> 
  filter(Username == users[focal_player_used]) |> 
  arrange(UTCDate, UTCTime) |> 
  slice(1:(n() - 1000)) |> 
  mutate(game = row_number(), glicko = ifelse(Username == White, WhiteElo, BlackElo)) %>%
  dplyr::select(game, glicko)
num_prev = nrow(true_glicko_first) #number of games before final simulated games

#true glicko rating of final 1000 games
true_glicko_final = focal_init_data %>%
  mutate(game = row_number() + num_prev, glicko = ifelse(focal_white == 1, WhiteElo, BlackElo)) %>%
  dplyr::select(game, glicko)

#glicko rating across all sims on final 100 games
final_sims = data.frame(glicko_sim, check.names = FALSE) %>%
  mutate(game = row_number() + num_prev) %>%
  pivot_longer(cols = -game, names_to = "Sim", values_to = "glicko")

#mean glicko rating for each game across all sims
sim_mean_glicko = final_sims %>% group_by(game) %>% summarise(mean_glicko = mean(glicko))

#plot
ppd_rating_plot = ggplot(data = true_glicko_first, aes(x = game, y = glicko)) +
  geom_line(linewidth = 0.5, col = "darkred") + 
  geom_line(data = final_sims, mapping = aes(group = Sim), alpha = 0.05) +
  geom_line(data = sim_mean_glicko, mapping = aes(x = game, y = mean_glicko), 
            alpha = 0.7) +
  geom_line(data = true_glicko_final, col = "red", linewidth = 0.5) +
  ylim(c(min(glicko_sim) - 50, max(glicko_sim) + 50)) +
  xlim(c(num_prev - 1500, num_prev + num_games)) +
  xlab("Games Played") + ylab("Glicko2 Rating") +
  theme_single_y()
ppd_rating_plot

# #save magnus
# ggsave(plot = ppd_rating_plot, 
#        filename = here("Paper_WriteUp", "paper_figures", "fig7a.png"), dpi = 1000,
#        height = 4.5, width = 10)
# #save nihal
# ggsave(plot = ppd_rating_plot, 
#        filename = here("Paper_WriteUp", "paper_figures", "fig7b.png"), dpi = 1000,
#        height = 4.5, width = 10)
```
